Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-12-09 paper/manuscript.tex

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[most]{tcolorbox}
\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[1]{Levi N. Naden}
\author[4*]{Samarjeet Prasad}
\author[5*]{Julia E. Rice}
\author[1,9]{Andrea Rizzi}
\author[2*]{Jenke Scheen}
\author[6*]{Michael Shirts}
\author[11]{Gary Tresadern}
\author[10*]{Huafeng Xu}

@@

\section{Abbreviations}
\label{sec:abbreviations}

\todo[inline]{JDC: Should we add an Abbreviations section? Maybe it can be in a shaded box?}

\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black]
    \begin{center}
        \textit{In alphabetical order:}
    \end{center}
    \textit{CPU} - Central Processing Unit\\
    \textit{FE} - Free Energy\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common SubStructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration

\end{tcolorbox}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig1_what_is_alchemy/Fig1.pdf}    \caption{Illustration of different types of free energies differences that can be computed using alchemical free energy calculations. A: Change in free energy due to a conformational change. B: Partition coefficient such as $\log P$ or $\log D$, here as an example the partition coefficient between methanol and water of a given molecule. C: Free energies of membrane insertions. D: Simulations of portein or other host molecule simulations. E: Absolute free energies of binding of a small molecule to a host (e.g. protein), F: Relative free energy of binding of one molecule with respec to another, here toluene and benzyl alcohol.
    \label{fig:fig1_what_is_alchemy}
    }
\end{figure}

@@

\subsection*{Free energies of hydration or partition coefficients}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are much generally much more straight forward. Using even something such as bable to generate a 3D minimised structure of the solute you would like to compute a free energy of hydration or partition coefficient can be sufficient for preparation of these types of simulations. However, in these cases a careful choice of forcefields, as well as water models or organic solvents is essential. See for example XX.et al~\cite{xxx} for a good discussion on these choices. 
\todo[inline, color=green!20]{Just a note as a reminder: Chirality!}
%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{What simulation protocol should I choose? -- Step 2}
\label{sec:step2}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}, which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
\subsection*{Absolute and relative free energy calculations have some differences}
Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.
\todo[inline, color=green!20]{JM: single/dual hybrid toplogies what are they? ASJSM: These have been addressed below}

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{Samar: write section}
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schrodinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations. The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{LiveCOMS-gromacs}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM sctipts for solvation and ligand-binding free energy calculations.
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

@@

\clearpage
\section{Benchmark Datasets}
\label{sec:benchmark}
\vspace{5mm}
\begin{tabular}{cccc}
\caption{Overview of benchmark datasets}
\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
\hline
\multicolumn{4}{|c|}{D3R Grand Challenges \cite{D3R}} \\
\hline
GC3 \cite{gaieb2019} & 6 & 266 & various \\
GC2 \cite{gaieb2018} & 1 & 102 & various \\
GC2015 \cite{gathiaka2016} & 2 & 215 & various \\
\hline
\multicolumn{4}{|c|}{SAMPL Challenges \cite{SAMPL}} \\
\hline
SAMPL6 \cite{rizzi2018} & 3 & 21 & various \\
SAMPL5 \cite{yin2017} & 3 & 22 & various \\
SAMPL4 \cite{muddana2014} & 2 & 23 & various \\
\hline
\multicolumn{4}{|c|}{Schrödinger Datasets} \\
\hline
FEP+ Dataset \cite{wang2015} & 8 & 199 & OPLS2.1 \\
FEP+ Dataset \cite{harder2016} & 8 & 199 & OPLS3 \\
FEP+ Dataset \cite{roos2019} & 8 & 199 & OPLS3e \\
FEP+ Dataset \cite{song2019} & 8 & 199 & AMBER18 \\
Fragment Optimization \cite{steinbrecher2015} & 8 & 96 & OPLS2.1 \\
Scaffold Hopping \cite{wang2017} & 6 & 21 & OPLS3 \\
Macrocycles \cite{yu2017} & 7 & 33 & OPLS3 \\
\hline
\multicolumn{4}{|c|}{Further Suggested Datasets} \\
\hline
Cucurbit[7]uril (CB7) \cite{mobley2017} & 1 & 15 & NA \\
Deep cavity cavitand \cite{mobley2017} & 2 & 19 & NA \\
T4 Lysozyme \cite{mobley2017} & 2 & 20 & NA \\
\hline
\label{tab:benchmarks}
\end{tabular}

\todo[inline, color={green!20}]{ASJSM: Please check whether you can format this in a more appealing way.}
Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al. \cite{williams2017}. For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf. the recently published study by Pérez-Benito et al. \cite{perez2019}.
An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson \cite{mobley2017} or on alchemistry.org \cite{alchemistry}. These include cyclodextrins, the Cytochrome c peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets \cite{paliwal2011}.

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{url}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[most]{tcolorbox}
\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[1]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[5]{Julia E. Rice}
\author[1,9]{Andrea Rizzi}
\author[2]{Jenke Scheen}
\author[6*]{Michael Shirts}
\author[11]{Gary Tresadern}
\author[10]{Huafeng Xu}

@@

%\label{sec:abbreviations}
\begin{tcolorbox}[title=Abbreviations, colback=blue!10!white]
    \textit{CPU} - Central Processing Unit\\
    \textit{FE} - Free Energy\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common SubStructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration
\end{tcolorbox}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig1_what_is_alchemy/Fig1.pdf}    \caption{Illustration of different types of free energies differences that can be computed using alchemical free energy calculations. A: Change in free energy due to a conformational change. B: Partition coefficient such as $\log P$ or $\log D$, here as an example the partition coefficient between methanol and water of a given molecule. C: Free energies of membrane insertions. D: Simulations of protein or other host molecule simulations. E: Absolute free energies of binding of a small molecule to a host (e.g. protein), F: Relative free energy of binding of one molecule with respec to another, here toluene and benzyl alcohol.
    \label{fig:fig1_what_is_alchemy}
    }
\end{figure}

@@

\subsection*{Free energies of hydration or partition coefficients}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are much generally much more straight forward. Using even something such as bable to generate a 3D minimised structure of the solute you would like to compute a free energy of hydration or partition coefficient can be sufficient for preparation of these types of simulations. However, in these cases a careful choice of forcefields, as well as water models or organic solvents is essential. See for example XX.et al~\cite{xxx} for a good discussion on these choices. 
\todo[inline, color=green!20]{Just a note as a reminder: Chirality!}
%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{What simulation protocol should I choose? -- Step 2}
\label{sec:step2}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}, which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
\subsection*{Absolute and relative free energy calculations have some differences}
Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{Samar: write section}
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schrodinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design. 
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations. The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{LiveCOMS-gromacs}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM sctipts for solvation and ligand-binding free energy calculations.
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

@@

\clearpage
\section{Benchmark Datasets}

\label{sec:benchmark}
\vspace{5mm}

\begin{tabular}{cccc}
\caption{Overview of benchmark datasets}
\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
\hline
\multicolumn{4}{|c|}{D3R Grand Challenges \cite{D3R2019Dec}} \\
\hline
GC3 \cite{gaieb2019} & 6 & 266 & various \\
GC2 \cite{gaieb2018} & 1 & 102 & various \\
GC2015 \cite{gathiaka2016} & 2 & 215 & various \\
\hline
\multicolumn{4}{|c|}{SAMPL Challenges \cite{SAMPL}} \\
\hline
SAMPL6 \cite{rizzi2018} & 3 & 21 & various \\
SAMPL5 \cite{yin2017} & 3 & 22 & various \\
SAMPL4 \cite{muddana2014} & 2 & 23 & various \\
\hline
\multicolumn{4}{|c|}{Schrödinger Datasets} \\
\hline
FEP+ Dataset \cite{wang2015} & 8 & 199 & OPLS2.1 \\
FEP+ Dataset \cite{harder2016} & 8 & 199 & OPLS3 \\
FEP+ Dataset \cite{roos2019} & 8 & 199 & OPLS3e \\
FEP+ Dataset \cite{song2019} & 8 & 199 & AMBER18 \\
Fragment Optimization \cite{steinbrecher2015} & 8 & 96 & OPLS2.1 \\
Scaffold Hopping \cite{wang2017} & 6 & 21 & OPLS3 \\
Macrocycles \cite{yu2017} & 7 & 33 & OPLS3 \\
\hline
\multicolumn{4}{|c|}{Further Suggested Datasets} \\
\hline
Cucurbit[7]uril (CB7) \cite{mobley2017} & 1 & 15 & NA \\
Deep cavity cavitand \cite{mobley2017} & 2 & 19 & NA \\
T4 Lysozyme \cite{mobley2017} & 2 & 20 & NA \\
Merck benchmarking set \cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\

\hline
\label{tab:benchmarks}
\end{tabular}

Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al. \cite{williams2017}. For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf. the recently published study by Pérez-Benito et al. \cite{perez2019}.
An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson \cite{mobley2017} or on alchemistry.org \cite{alchemistry}. These include cyclodextrins, the Cytochrome c peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets \cite{paliwal2011}.
